---
title: "English Premier League Stats (2021-2025)"
author: "Mike May, Jessica Vescera, Melina Rodriguez, Frank Zawacki"
date: "2026-03-24"
output: html_document
---
Introduction:
This project explores match performance in the English Premier League from 2021 to 2024. The goal is to analyze home field advantage, team performance metrics, and match outcomes using descriptive statistics and visualizations.




##Load Data
```{r setup, include=FALSE}
library(readr)
data <- read.csv("Documents/mydata.csv", header= FALSE, stringsAsFactors= FALSE)
head(data)
data
```

##Fix Column Names
```{r}

#Use first row as column names
colnames(data) <- data[1, ]

#Remove first row from dataset 
data <- data[-1, ]

#Reset row names 
rownames(data) <- NULL

# Clean column names 
names(data) <- make.names(names(data))

# Check updated structure
str(data)
```

#Convert Data Types
```{r}
# Convert key columns to numeric
data$Goals.Home <- as.numeric(data$Goals.Home)
data$Away.Goals <- as.numeric(data$Away.Goals)
data$home_shots <- as.numeric(data$home_shots)
data$away_shots <- as.numeric(data$away_shots)
data$home_possessions <- as.numeric(data$home_possessions)
data$away_possessions <- as.numeric(data$away_possessions)

# Fix attendance (remove commas, then convert)
data$attendance <- as.numeric(gsub(",", "", data$attendance))
```

##Create Match Outcome Variables


```{r pressure, echo=FALSE}
#Home Win, Away Win, Draw
data$result <- ifelse(data$Goals.Home > data$Away.Goals, "Home Win",
                      ifelse(data$Goals.Home < data$Away.Goals, "Away Win", "Draw"))

table(data$result)
```

##Home Vs Away Win Percentage
```{r}
# Count results
result_counts <- table(data$result)

# Convert to percentages
result_percent <- prop.table(result_counts) * 100

# Print results
print(result_counts)
print(result_percent)
```
```{r}
barplot(table(data$result),
        main = "Match Outcomes",
        col = c("green", "red", "gray"))
```
Interpretation:

The bar chart shows the distribution of match outcomes across the dataset. Home wins appear to occur more frequently than away wins and draws, which suggests a potential home-field advantage in the Premier League during this period.
Away wins are still fairly common, indicating that away teams remain competitive, but they do not occur as often as home wins. Draws are the least frequent outcome, showing that most matches tend to produce a clear winner rather than ending evenly.
Overall, this pattern supports the idea that playing at home may provide a slight advantage, but matches are still relatively competitive across all outcomes.


##Goal Analysis 
```{r}
mean(data$Goals.Home, na.rm = TRUE)
mean(data$Away.Goals, na.rm = TRUE)
```

```{r}
boxplot(data$Goals.Home, data$Away.Goals,
        names = c("Home Goals", "Away Goals"),
        main = "Goals Scored: Home vs Away",
        ylab = "Goals")
```


Interpretation:

This boxplot compares goals scored at home vs away.
Home and away teams look pretty similar overall. The median number of goals is close for both, but home teams appear to score slightly more on average. The spread (IQR) is also similar, meaning the variability in goals is about the same in both cases.
Both groups have a few outliers, especially on the higher end, showing that in some matches teams scored a lot more goals than usual. However, these are rare and don’t change the overall pattern much.
Overall, the chart suggests a slight home advantage, but not a huge difference between home and away goal scoring.

##Shots vs Goals
```{r}
plot(data$home_shots, data$Goals.Home,
     main = "Home Shots vs Goals",
     xlab = "Shots",
     ylab = "Goals")
```


Interpretation:
There’s a general upward trend showing that as home teams take more shots, they tend to score more goals. However, the spread of points shows that more shots don’t always guarantee more goals, so efficiency still matters.

```{r}
plot(data$away_shots, data$Away.Goals,
     main = "Away Shots vs Goals",
     xlab = "Shots",
     ylab = "Goals")
```


Interpretation:
Away teams also show a positive relationship between shots and goals, but the pattern looks slightly more spread out. This suggests that away teams may be a bit less consistent in converting chances compared to home teams.

##Possession Analysis
```{r}
boxplot(home_possessions ~ result, data = data,
        main = "Home Possession by Match Result",
        xlab = "Match Result",
        ylab = "Home Possession (%)")
```


Interpretation:
Home teams tend to have higher possession in games they win compared to games they lose. However, the overlap between groups shows that possession alone doesn’t determine the outcome, and other factors like finishing and defense also play a role.

##Random Forest Model


```{r}
# Load packages
library(randomForest)

# Create binary outcome (Home Win = 1, else 0)
data$home_win_binary <- ifelse(data$result == "Home Win", 1, 0)
data$home_win_binary <- as.factor(data$home_win_binary)

# Select features
features <- data[, c("home_possessions",
                     "away_possessions",
                     "home_shots",
                     "away_shots",
                     "home_fouls",
                     "away_fouls",
                     "home_corners",
                     "away_corners",
                     "attendance")]

# Convert to numeric where needed
features$home_possessions <- as.numeric(features$home_possessions)
features$away_possessions <- as.numeric(features$away_possessions)
features$home_shots <- as.numeric(features$home_shots)
features$away_shots <- as.numeric(features$away_shots)
features$home_fouls <- as.numeric(features$home_fouls)
features$away_fouls <- as.numeric(features$away_fouls)
features$home_corners <- as.numeric(features$home_corners)
features$away_corners <- as.numeric(features$away_corners)
features$attendance <- as.numeric(gsub(",", "", features$attendance))

# Remove missing values
model_data <- na.omit(cbind(features, home_win_binary = data$home_win_binary))

# Split into X and y
X <- model_data[, -ncol(model_data)]
y <- model_data$home_win_binary

# Train/test split (80/20)
set.seed(42)
train_index <- sample(1:nrow(model_data), 0.8 * nrow(model_data))

X_train <- X[train_index, ]
X_test  <- X[-train_index, ]
y_train <- y[train_index]
y_test  <- y[-train_index]

# Train Random Forest model
rf_model <- randomForest(x = X_train, y = y_train,
                         ntree = 100)

# Predictions
y_pred <- predict(rf_model, X_test)

# Accuracy
accuracy <- sum(y_pred == y_test) / length(y_test)
print(paste("Accuracy:", round(accuracy, 3)))

# Confusion Matrix
cm <- table(Predicted = y_pred, Actual = y_test)
print("Confusion Matrix:")
print(cm)

# Feature Importance
importance_vals <- importance(rf_model)
print("Feature Importance:")
print(importance_vals)

# Plot Feature Importance
varImpPlot(rf_model)


# Get importance values
importance_vals <- importance(rf_model)

# Convert to data frame
importance_df <- data.frame(
  feature = rownames(importance_vals),
  importance = importance_vals[,1]
)

# Order by importance (largest first)
importance_df <- importance_df[order(importance_df$importance, decreasing = TRUE), ]

# Bar plot
barplot(importance_df$importance,
        names.arg = importance_df$feature,
        las = 2,  # makes labels vertical
        col = "steelblue",
        main = "Feature Importance - Random Forest",
        ylab = "Importance")



library(ggplot2)

cm_df <- as.data.frame(cm)

ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq)) +
  labs(title = "Confusion Matrix - Random Forest") +
  theme_minimal()
```

Random Forest Interpretation:
The random forest results show that variables like attendance and shots, particularly away shots, are among the most important in predicting whether the home team wins. Possession also contributes to the model, while fouls and corners appear to have less influence. This suggests that attacking activity and overall game intensity play a larger role in determining match outcomes.

The confusion matrix indicates that the model has moderate predictive performance. While it correctly classifies some outcomes, there are still a number of misclassifications. In particular, the model tends to predict “no home win” more frequently, which leads to missed home win predictions.

Overall, the model captures some meaningful patterns in the data, but its accuracy is limited. This suggests that adding more relevant features or further refining the model could help improve its performance.